---
title: "ConScapeRtools User Guide"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    fig_width: 5.75
    fig_height: 4.25
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{ConScapeRtools User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
library(ConScapeRtools)
```

# Background

Landscape connectivity is usually shaped by two related pieces of information:
where habitat is available, and how easily movement can occur between habitat
locations. `ConScapeRtools` is an R interface to the Julia package
`ConScape`, with an emphasis on large raster landscapes that are too large to
solve comfortably as a single graph.

The main workflow is:

1. Choose or calibrate movement and decay parameters with `tile_design()`.
2. Tile source quality, target quality, and movement affinity rasters with
   `conscape_prep()`.
3. Run ConScape on each tile with `run_conscape()`.
4. Reassemble tile-level outputs with `mosaic_conscape()`.

The package keeps older argument names for compatibility, but the underlying
ConScape model is easiest to understand using ConScape's own vocabulary.

```{r api-slots}
conscape_api_slots()
```

In brief:

- `target_qualities` identify cells that can receive connectivity.
- `source_qualities` identify cells that generate connectivity.
- `affinities` describe movement permeability or movement likelihood.
- `cost_function` describes how graph movement costs are derived from
  affinities.
- `theta` controls the randomized shortest path model.
- `distance_scale` controls the exponential transformation from distance to
  proximity.

# Data

This guide uses the small example rasters installed with the package. In a real
analysis, use projected rasters with map units that match the movement distance
you want to interpret.

```{r read-data, eval=FALSE}
library(terra)
library(ConScapeRtools)

hab_file <- system.file("extdata", "suitability.asc", package = "ConScapeRtools")
aff_file <- system.file("extdata", "affinity.asc", package = "ConScapeRtools")

habitat <- terra::rast(hab_file)
affinity <- terra::rast(aff_file)

plot(c(habitat, affinity))
```

The movement raster should be an affinity or permeability surface, where larger
values mean easier movement. If your starting layer is resistance, transform it
before using it as `r_mov` or `affinities`.

# Tile Design

Before tiling a large landscape, it is useful to estimate a distance decay
parameter and an explicit tile overlap. `tile_design()` builds a fully connected
synthetic ideal-habitat ConScape problem, computes expected costs from the
center cell, and chooses an exponential distance scale so that proximity has
decayed to a small threshold at the specified maximum distance.
The calibration uses fully connected movement with single center source and
target cells, so ConScape returns one center-referenced expected-cost column
instead of an expensive all-targets matrix.

The overlap is controlled by `trim_threshold`: the returned `tile_trim` is the
landmark-rounded distance at which the calibrated proximity is no greater than
that threshold. The interior tile width is chosen separately. You can provide a
specific `tile_width`, an approximate `max_tile_cells` limit, or let the default
`overlap_area_factor` choose a width that balances retained tile area against
overlap overhead.
`distance_scale` is a ConScape expected-cost scale, not a map distance. The
trim distance is therefore read from the calibrated cost-to-proximity curve in
map units before it is rounded to the raster and landmark grid.

```{r tile-design, eval=FALSE}
jl_home <- "C:/Users/you/AppData/Local/Programs/Julia-1.12.6/bin"

td <- tile_design(
  r_mov    = affinity,
  r_target = habitat,
  max_d    = 8500,
  theta    = 0.1,
  jl_home  = jl_home,
  trim_threshold = 0.01,
  overlap_area_factor = 2,
  landmark = 5L
)

td
td$diagnostics
```

The returned list includes `distance_scale` (the calibrated decay parameter),
`tile_d`, `tile_trim`, and `theta` - all in the map units of the input raster.
Pass `tile_d`, `tile_trim`, and the same `landmark` to `conscape_prep()`, and
`distance_scale` and `theta` to `run_conscape()`. The `diagnostics` element
records the requested trim, effective trim after landmark rounding, expected
tile count, overlap area factor, and proximity at the final trim distance.

# Preparing Tiles

`conscape_prep()` writes matching ASCII tiles for target qualities, source
qualities, and movement affinities. All rasters must have the same extent and
resolution.

```{r prep, eval=FALSE}
prep <- conscape_prep(
  tile_d = td$tile_d,
  tile_trim = td$tile_trim,
  asc_dir = "conscape_tiles",
  r_target = habitat,
  r_src = habitat,
  r_mov = affinity,
  target_threshold = 0,
  landmark = td$landmark,
  clear_dir = TRUE
)

prep$input_slots
prep$target_mode
```

The `landmark` argument is ConScape's coarse-graining window size. Larger
values can make large runs faster because target qualities are aggregated to
fewer target cells. Tile dimensions and overlaps are aligned to this value so
the coarse-grained target cells fall in consistent locations across tiles.

## Target mode: how each tile's RSP problem is posed

`conscape_prep()` exposes a `target_mode` argument. The default,
`target_mode = "auto"`, resolves to `"center"` whenever `tile_d`/`tile_trim`
(or `centersize`/`buffer`) are supplied. The choice between `"center"` and
`"full"` controls which cells in a buffered tile act as RSP absorbing targets
when ConScape solves the per-tile linear system
`Z = (I - W) \ I_target`.

- `"center"` *(default)* restricts the target raster written for each tile
  to its **center** cells (the interior region defined by `tile_d`). Source
  qualities and affinities are kept in full across the buffer. Adjacent
  tiles' centers are disjoint and together cover the landscape exactly once,
  so the tile-level outputs are combined with `mosaic_conscape(method = "sum")`
  and the result equals the untiled solution up to buffer-truncation error.
  This is the mathematically correct windowing of the ConScape problem and
  is the same per-window math that ConScape's experimental dev
  `WindowedProblem` uses, executed against the stable bundled Julia API.

- `"full"` keeps every cell of the buffered tile as a target. Each tile
  produces a full surface and overlapping tiles are averaged with
  `mosaic_conscape(method = "mosaic")` (the legacy default). The averaged
  estimator is **biased**: each tile undercounts targets that fall in other
  tiles. Use `"full"` only to reproduce results from earlier ConScapeRtools
  versions; expect smoothing artifacts near tile borders.

`run_conscape()` reads `prep$target_mode` and selects the matching mosaic
reduction automatically, so the two defaults stay aligned without further
configuration.

### Convergence behavior in practice

Both `"center"` (sum mosaic) and `"full"` (mean mosaic) tile modes converge
to the untiled solution as buffer grows, by two different mechanisms:

* `"center"` + sum mosaic: each tile restricts targets to its center cells
  and sums per-cell contributions across overlapping tiles. The reconstruction
  is exact up to paths that exit every relevant tile's buffer; that
  truncation is bounded by proximity at the buffer edge, so error decays
  approximately as `exp(-buffer / distance_scale)`.

* `"full"` + mean mosaic: each tile keeps full buffered targets and computes
  a per-cell fcon biased low because it cannot see targets in other tiles.
  As buffer grows toward the landscape size, each tile's view approaches the
  full untiled landscape and the bias shrinks; the mean of many near-correct
  per-tile estimates approaches the untiled value.

**Why `"center"` is still the default.** The two modes' per-cell errors are
comparable at small buffer. The decisive difference is per-tile compute:
each tile in `"center"` mode solves `Z = (I - W) \ I_target` over only its
center-cell target columns, so per-tile cost stays bounded as the landscape
grows. `"full"` mode requires every tile to solve `Z` over its full
buffered target set, which becomes prohibitive on landscapes too large to
fit in a single tile's buffered window. If a landscape *is* small enough
that every tile's buffered window covers most of it, you should typically
not be tiling at all.

A reproducible convergence comparison on the package data lives at
`tests/testthat/test-tiled-untiled-convergence.R`. With `centersize = 20`,
`distance_scale = 20`, `theta = 0.1` on an 80x80 crop of the package
suitability/affinity rasters, the test reports (max-abs error against the
untiled reference, both modes, buffer values 5 to 40):

```
       mode buffer   btwn      fcon
 center_sum      5  343.27     7.57
  full_mean      5  320.97     6.46
 center_sum     10  190.73     3.55
  full_mean     10  160.08     3.17
 center_sum     20   40.58     0.58
  full_mean     20   24.78     0.59
 center_sum     40    1.74     0.024
  full_mean     40    0.97     0.024
```

Both modes converge to near-zero error at large buffer; `"center"` is the
scalable choice.

# Running Connectivity Metrics

The default `run_conscape()` call computes two connectivity surfaces:

- `btwn`: ConScape `betweenness_kweighted()` — how strongly each cell
  contributes to quality-weighted movement across the landscape.
- `fcon`: ConScape `connected_habitat()` — how much quality-weighted target
  habitat is effectively reachable from each source cell.

```{r run-default, eval=FALSE}
cs <- run_conscape(
  conscape_prep  = prep,
  out_dir        = "conscape_results",
  theta          = td$theta,
  distance_scale = td$distance_scale,
  jl_home        = jl_home,
  parallel       = TRUE,
  workers        = 4
)

plot(cs)
```

For small to moderate rasters that do not require tiling, pass `SpatRaster`
objects directly using the ConScape-aligned argument names.

```{r run-single, eval=FALSE}
single <- run_conscape(
  out_dir          = "single_results",
  target_qualities = habitat,
  source_qualities = habitat,
  affinities       = affinity,
  theta            = td$theta,
  distance_scale   = td$distance_scale,
  landmark         = 5L,
  jl_home          = jl_home
)
```

Additional metrics can be requested with `metrics`.

```{r metrics, eval=FALSE}
cs_more <- run_conscape(
  conscape_prep = prep,
  out_dir = "conscape_more_metrics",
  theta = td$theta,
  distance_scale = td$distance_scale,
  metrics = c("connected_habitat", "betweenness_kweighted",
              "betweenness_qweighted", "criticality"),
  jl_home = jl_home
)
```

# Sensitivity Analysis

Sensitivity analysis asks how much a landscape-level connectivity summary would
change if a local landscape component changed. The current ConScape sensitivity
framework supports perturbations with respect to:

- `"Q"`: habitat quality.
- `"A"`: affinity or permeability.
- `"C"`: movement cost.
- `"A&C=f(A)"`: linked affinity and cost when cost is a function of affinity.
- `"C&A=f(C)"`: linked cost and affinity when affinity is treated as a function
  of cost.

Use `conscape_sensitivity()` to describe the requested sensitivity outputs.

```{r sensitivity-spec}
sens <- conscape_sensitivity(
  wrt = c("Q", "A&C=f(A)"),
  landscape_measure = "sum",
  unitless = TRUE
)

sens
```

With `unitless = TRUE`, outputs are elasticities. These are often easier to
compare across surfaces because they describe response to proportional change.
Use `unitless = FALSE` for raw unit-change sensitivities.

```{r sensitivity-run, eval=FALSE}
prep_sens <- conscape_prep(
  tile_d = td$tile_d,
  tile_trim = td$tile_trim,
  asc_dir = "conscape_sensitivity_tiles",
  r_target = habitat,
  r_src = habitat,
  r_mov = affinity,
  target_threshold = 0,
  landmark = 1L,
  clear_dir = TRUE
)

cs_sens <- run_conscape(
  conscape_prep  = prep_sens,
  out_dir        = "conscape_sensitivity",
  theta          = td$theta,
  distance_scale = td$distance_scale,
  metrics        = c("connected_habitat", "betweenness_kweighted"),
  sensitivity    = sens,
  jl_home        = jl_home
)

names(cs_sens)
plot(cs_sens, layers = c("elasticity_quality",
                         "elasticity_affinity_cost_linked"))
```

There are two important practical constraints.

First, the current ConScape sensitivity implementation assumes source and target
qualities are equal. Because `landmark > 1L` coarse-grains target qualities,
`run_conscape()` requires `landmark = 1L` by default when sensitivity is
requested. When using `conscape_prep`, this means preparing the sensitivity
tiles with `landmark = 1L`. This keeps the target-quality surface aligned with
the source-quality surface.

Second, tiled sensitivity is a tile-local approximation. It uses the same tile
overlap and mosaicking strategy as `btwn` and `fcon`, but sensitivity to
movement affinities and costs can be more boundary-sensitive because connector
cells and edges near tile borders may influence interior paths. Use conservative
tile overlap, inspect edge artifacts, and prefer `landscape_measure = "sum"` for
tiled sensitivity workflows.

# Tiled Sensitivity Recommendations

A conservative tiled sensitivity workflow is:

```{r tiled-sensitivity-template, eval=FALSE}
td <- tile_design(
  r_mov    = affinity,
  r_target = habitat,
  max_d    = 8500,
  theta    = 0.1,
  jl_home  = jl_home
)

prep_sens <- conscape_prep(
  tile_d           = td$tile_d,
  tile_trim        = td$tile_trim,
  asc_dir          = "conscape_sensitivity_tiles",
  r_target         = habitat,
  r_src            = habitat,
  r_mov            = affinity,
  target_threshold = 0,
  landmark         = 1L,
  clear_dir        = TRUE
)

cs_sens <- run_conscape(
  conscape_prep         = prep_sens,
  out_dir               = "conscape_sensitivity_results",
  theta                 = td$theta,
  distance_scale        = td$distance_scale,
  landmark              = 1L,
  cost_function         = "minuslog",
  connectivity_function = "expected_cost",
  sensitivity           = conscape_sensitivity(
    wrt               = c("Q", "A&C=f(A)"),
    landscape_measure = "sum",
    unitless          = TRUE
  ),
  jl_home  = jl_home,
  parallel = TRUE,
  workers  = 4
)
```

For full-landscape sensitivity without tiling, pass the rasters directly to
`run_conscape()`. This avoids tile-boundary approximation, but can be much more
computationally demanding.

```{r full-sensitivity, eval=FALSE}
full_sens <- run_conscape(
  out_dir          = "full_sensitivity_results",
  target_qualities = habitat,
  source_qualities = habitat,
  affinities       = affinity,
  theta            = td$theta,
  distance_scale   = td$distance_scale,
  landmark         = 1L,
  metrics          = "connected_habitat",
  sensitivity      = conscape_sensitivity(wrt = c("Q", "A&C=f(A)")),
  jl_home          = jl_home
)
```

# Output Interpretation

The default metric outputs answer different ecological questions.

- `fcon` / `connected_habitat`: how much quality-weighted target habitat is
  effectively connected to each source cell after distance has been transformed
  to proximity.
- `btwn` / `betweenness_kweighted`: how strongly each cell contributes to
  quality-weighted movement among connected source-target pairs.
- `elasticity_quality`: expected proportional change in the landscape summary
  following proportional change in habitat quality at a cell.
- `elasticity_affinity_cost_linked`: expected proportional change in the
  landscape summary following proportional change in affinity when cost changes
  consistently with the selected `cost_function`.

High sensitivity does not automatically mean a cell has high current
connectivity. It means the chosen landscape summary is responsive to change at
that location. Sensitivity maps are therefore useful for comparing restoration,
degradation, or management scenarios.

# Other Functions

`mosaic_conscape()` can be used directly when tile outputs already exist.

```{r direct-mosaic, eval=FALSE}
btwn <- mosaic_conscape(
  out_dir = file.path("conscape_results", "btwn"),
  mask = terra::rast(file.path(prep$asc_dir, "mask", "mask.asc")),
  tile_trim = prep$tile_trim,
  method = "merge"
)
```

The plotting method understands all raster outputs stored in a
`ConScapeResults` object.

```{r plot-layers, eval=FALSE}
plot(cs_sens)
plot(cs_sens, layers = c("fcon", "elasticity_quality"))
```

# Practical Checklist

Before running a large analysis:

1. Confirm `affinities` / `r_mov` is a permeability surface, not resistance.
   Transform resistance to affinity before passing it to ConScapeRtools.
2. Run `tile_design()` to obtain a landscape-calibrated `distance_scale`,
   `theta`, `tile_d`, `tile_trim`, `landmark`, and diagnostics. Check
   `td$diagnostics$overlap_area_factor` and
   `td$diagnostics$proximity_at_effective_trim` before launching a large run.
3. Before launching, run `window_design_scenarios(r, design = td)` to
   see how per-window peak RAM, window count, and total workload would
   change at alternative `centersize` values. Pass `target_ram_gb` and
   `workers` to flag the largest `centersize` that fits your machine's
   memory budget; adjusting `centersize` does not change the
   mathematical answer, only the per-window memory footprint and
   throughput. See the "Tile sizing: RAM, throughput, and tradeoffs"
   subsection of `vignette("Tiled_ConScape_Validation_and_Performance",
   package = "ConScapeRtools")`.
4. Choose `theta` based on the movement model: small values (near 0) allow more
   random movement; large values approach deterministic least-cost paths.
5. Leave `target_mode` at its default. `"auto"` resolves to `"center"` for any
   tiled run, which restricts each tile's RSP absorbing target set to the tile
   center and combines outputs with a sum mosaic. This converges to the untiled
   solution as the buffer grows. Pass `target_mode = "full"` only when
   reproducing older results that depended on the legacy mean-mosaic estimator.
6. Use `landmark > 1L` for faster default metric runs when coarse-grained target
   qualities are acceptable.
7. Use `landmark = 1L` for sensitivity workflows to keep target and source
   qualities identical.
8. Start with the default `metrics` (`connected_habitat` and
   `betweenness_kweighted`) before adding additional metrics or sensitivity.
9. Add sensitivity outputs only after the base workflow is verified and stable.
10. Compare tiled and untiled runs on a small landscape crop before scaling up.
    The convergence test (`tests/testthat/test-tiled-untiled-convergence.R`)
    demonstrates the expected error behavior of each tile mode as a function of
    buffer width and serves as a reproducible diagnostic template.
